---
title: "Distributions"
teaching: 42
exercises: 47
questions: 
- "FIXME"

objectives:
- "FIXME"

keypoints:
- "FIXME"
source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
```

We have some data on the mass of penguins. We expect some penguins to weigh 
more than others. Perhaps it is a smaller kind of penguin. Perhaps it is a 
female penguin, and they are smaller than male penguins. And maybe some penguins
are fatter than other penguins.

Ranking penguins from light to heavy, 
how many are there, or, what is the distribution of their weight?

## Histograms

### What are they?

Histograms takes all the weights of the penguins, divides them into intervals, 
or bins, of weight, eg one bin with 3300 to 3400 grams, and the next bin from 
3400 to 3500 grams. Then we count how many penguins are in a specific bin. 
And plot it. It might look like this:
```{r echo=F}
penguins %>% ggplot(aes(body_mass_g)) + geom_histogram()
```


### What do we use them for?

We typically use histograms to visualise the distribution of a variable. Is it
normal, bimodal, uniform or skewed? It can quickly reveal mistakes in data or 
weird outliers.

We can also use histograms to compare different variables. But beware of comparing
too many, it can make the graphs difficult to read, and the aim of visualising 
is to make things easier to understand.

### how do we make them?

The geom_ we are using here is geom_histogram(). It takes only one variable 
in the mapping.

```{r}
# library
library(ggplot2)
library(palmerpenguins)

# dataset:
data <-  penguins

# basic histogram
ggplot(data, aes(x=body_mass_g)) + 
  geom_histogram()
```
Built into the geom_histogram is the statistical transformation, that counts the 
number of observations in each bin.

Note the warning that this statistical transformation as default uses bins = 30.
This is the default number of bins. And it is almost quaranteed that it is not
the best number of bins.

Bins can be determined in two ways, either by providing a number of bins:
*geom_histogram(bins = 30)* 
or by specifying the width of the bins:
*geom_histogram(binwidth = 500)*


### Interesting variations

#### more than one on same axis
```{r}
penguins %>%
  filter(!is.na(sex)) %>% 
  ggplot( aes(x=body_mass_g, fill=sex)) +
  geom_histogram(alpha = 0.6, position = "identity")
```

Note that we do not map sex to color, but rather to fill. Color is the 
color of the outline of the individual bars, fill the inside of the 
bars.

#### Upside-down

Or mirrored histogram.

det skal vist gøres lidt mere læseligt...
```{r}
ggplot() +
  geom_histogram(aes(x = penguins %>% filter(sex == "female") %>% pull(body_mass_g), y = ..density.. ), fill="#69b3a2" ) +
  geom_histogram(aes(x = penguins %>% filter(sex == "male") %>% pull(body_mass_g), y = -..density.. ), fill= "#404080")

```


#### Grid

#### margin


### Think about

The number of bins (or their width, they are equivalent) can lead to very
different conclusions. Try several sizes.
```{r echo = F}
library(ggplot2)
library(tidyverse)
library(palmerpenguins)
library(cowplot)



# dataset:
data <-  penguins
bins <- c(15,30,45, 60)
# basic histogram
ggplot(data, aes(x=body_mass_g)) + 
  geom_histogram()

p1 <- ggplot(data, aes(x=body_mass_g)) + 
  geom_histogram(bins = bins[1]) +
  ggtitle(paste0("Bin number = ", bins[1]))


p2 <- ggplot(data, aes(x=body_mass_g)) + 
  geom_histogram(bins = bins[2]) +
  ggtitle(paste0("Bin number = ", bins[2]))


p3 <- ggplot(data, aes(x=body_mass_g)) + 
  geom_histogram(bins = bins[3]) +
  ggtitle(paste0("Bin number = ", bins[3]))

p4 <- ggplot(data, aes(x=body_mass_g)) + 
  geom_histogram(bins = bins[4]) +
  ggtitle(paste0("Bin number = ", bins[4]))



cowplot::plot_grid(p1, p2, p3, p4)

```


Weird and complicated color schemes does not add insight. Avoid them.

This is not a barplot! Histograms plot the distribution of a single variable.

Avoid comparing more than two, maybe three groups in the same histogram. 

Do not use unequal bin widths. 


## Boxplots

### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

### Think about

## Density

### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

### Think about

## Ridgeline



### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

### Think about

## Violin

### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

### Think about




{% include links.md %}