---
title: "Correlations"
teaching: 42
exercises: 47
questions: 
- "FIXME"

objectives:
- "FIXME"

keypoints:
- "FIXME"
source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
library(tidyverse)
library(palmerpenguins)
library(GGally)
library(ggExtra)
library(patchwork)
```

# Correlations

If one variable goes up, what happens to the other variable? How are they
correlated?

## Scatterplots

### What are they?

Shows the relation between two numeric variables. Each dot represents one 
observation. The position of the dot on the X-axis (horisontal, AKA abscissa), 
represents the value of the first variable for that observation. The position
of the dot on the Y-axis (vertical, AKA ordinate), represents the value of 
the second variable for that observation.

```{r scatter-what, echo = F}
ggplot(penguins, aes(x=body_mass_g, y=flipper_length_mm)) +
  geom_point()
```




### What do we use them for?

Typically used to show the relation between two variables. 

### how do we make them?

The geom_point() function makes the scatterplot. We need to provide the
mapping of two variables:
```{r scatter-how}
ggplot(penguins, aes(x=body_mass_g, y=flipper_length_mm)) +
  geom_point()
```


### Interesting variations

#### all combinations 

Since scatterplots provides a quick way of visualizing the correlation between
two variables, it can be useful to visualize all combinations of two variables
in our data.

Base-R does it like this:

```{r scatter-matrix}
plot(penguins)
```

The package `GGally` provides a function `ggpairs` that does this in the 
ggplot2 universe, making it easy to adjust the plot subsequently.

```{r scatter-matrix-ggally}
ggpairs(penguins)
```

Be careful - the plot can get very busy!

#### coloring

```{r scatter-how}
ggplot(penguins, aes(x=body_mass_g, y=flipper_length_mm, color = species)) +
  geom_point()
```






#### marginal distibution 

Scatterplots kan udvides med plots på margenen: Det er ggmarginal fra
ggextra der skal på banen hvis det skal være let.

```{r scatter_marginal_histogram}

p <- penguins %>% 
  ggplot(aes(flipper_length_mm, body_mass_g)) +
  geom_point() 
ggMarginal(p, type = "histogram")
```

Bemærk at det ggmarginal element der kommer ud af det, ikke er helt let at 
arbejde videre med. Pak det ind i wrap_elements() fra patchwork pakken, så 
kører det.

Der er yderligere muligheder:

```{r scatter_marginal_flere, echo = F}

marg_dens <- ggMarginal(p, type = "density") %>% wrap_elements() + ggtitle('type = "density"')
marg_violin <- ggMarginal(p, type = "violin") %>% wrap_elements() + ggtitle('type = "violin"')
marg_densigram <- ggMarginal(p, type = "densigram") %>% wrap_elements() + ggtitle('type = "densigram"')
marg_boxplot <- ggMarginal(p, type = "boxplot") %>% wrap_elements() + ggtitle('type = "boxplot"')

marg_dens + marg_violin + marg_densigram + marg_boxplot

```


### Think about


Overlapping points




## Connected scatter

### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

### Think about

## heatmap

### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations


## Correlogram


### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

## Bubble


### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

## Density 2D

### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

{% include links.md %}