---
title: "Ranking"
teaching: 42
exercises: 47
questions: 
- "FIXME"

objectives:
- "FIXME"

keypoints:
- "FIXME"
source: Rmd
---
tag et kig på:
https://clauswilke.com/dataviz/directory-of-visualizations.html

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("04-")
library(tidyverse)
library(patchwork)
```

Plot types useful for showing rankings - that is, which type of observation
is the most common, and which is the least common.


## Barplots

### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

### Think about
Be careful! It is tempting to use barplots for other stuff, like the 
mean value of two groups. This can be missleading. Here is an example:

```{r}
set.seed(47)
group1 <- rnorm(n = 100, mean = 1, sd = 1)
group2 <- rlnorm(n = 100, 
                 meanlog = log(1^2/sqrt(1^2 + 1^2)), 
                 sdlog = sqrt(log(1+(1^2/1^2))))
groups_long <- cbind(
  group1,
  group2
) %>% 
  as.data.frame() %>% 
  gather("group", "response", 1:2)

bar <- groups_long %>% 
  ggplot(aes(x = group, y = response)) +
  geom_bar(stat = "summary", fun = mean, 
           width = 0.7, alpha = 0.8,
           aes(fill = group)) + 
  stat_summary(geom = "errorbar", fun.data = "mean_se",
               width = 0.1, size = 1) 

dotplot <- groups_long %>% 
  ggplot(aes(x = group, y = response)) +
  ggbeeswarm::geom_quasirandom(
    shape = 21, color = "white", 
    alpha = 0.8, size = 3,
    aes(fill = group)
  ) +
  labs(x = "Group",
       y = "Response",
        caption = paste0("OH!!!\nP = ", 
                        signif(ks.test(group1, group2)$p.value, 2),
                        " (Kolmogorov–Smirnov test)")) +
  theme_classic() +
  theme(
    text = element_text(size = 12, face = "bold", color = "black"),
    axis.text = element_text(color = "black"),
    legend.position = "none",
    plot.title = element_text(size = 10),
    plot.caption = element_text(hjust = 0)
  ) +
  ggtitle(
    paste0(
      "group1: median = ", signif(median(group1), 2), 
      "; IQR = ", signif(IQR(group1), 2), "\n",
      "group2: median = ", signif(median(group2), 2), 
      "; IQR = ", signif(IQR(group2), 2)
    )
  )

wrap_plots(
  bar, dotplot, nrow = 1
)

```


## spider/radar plots

### What are they?

A two-dimensional chart designed to plot one or more series of values
over multiple quantitative variables.

### What do we use them for?

### how do we make them?

ggradar



### Interesting variations

### Think about
Det her er også noget skrammel...

https://www.data-to-viz.com/caveat/spider.html

## Wordclouds

### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

### Think about
Wordclouds are very popular.

But they have a lot of problems.

* They do not (neccesarily) group together words with the same meaning. "Difficult"
and "hard" might mean the same thing, but a standard wordcloud will treat them as two 
different words.
* They do not capture complex themes. "Expensive" is the same as "not cheap" or
"costs too much". This is subtly different from the previous point.
* They lack context. If something is helpful - what is helpful?
* They are prone to bias. Try to show people the same word cloud, and 
ask them to list the top five information in them.
* They obscure the relative importance of stuff. Is the first issue actually
as important as the second? Are the first three equally important?



## Parallel

### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

### Think about


## Lollipop

### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

### Think about


## Circular barplot

### What are they?

### What do we use them for?

### how do we make them?

### Interesting variations

### Think about




{% include links.md %}