---
title: "Tables"
teaching: 0
exercises: 0
questions: 
- "What are histograms?"
- "What are violin plots?"
- "What are density plots?"
- "What are boxplots?"

objectives:
- "FIXME"

keypoints:
- "FIXME"
source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("11-")
library(tidyverse)
library(patchwork)
library(palmerpenguins)
```

We do not typically think of tables as visualisations. But
they are often the best way of showing data. 

Charts use an abstraction to show the data. And sometimes that
abstraction gets in the way. Tables present data as close to raw
as possible. 

Use tables when data cannot be presented visually easily, or when
the data requires more specific attention.

The goal of presenting data is to make trends, connections, differences or similarities more apparent to the reader. 

```{r}

```


Maybe not a visualisation as we usually understand it.

But still

  https://themockup.blog/posts/2020-09-04-10-table-rules-in-r/
  
  
  Også denne, men den hører hjemme et andet sted.

{% include links.md %}